---
title             : "Lab8 Papaja"
shorttitle        : "lab8"

author: 
  - name          : "Yeaju Diana Kim"
    affiliation   : "1"
    corresponding : yes
    address       : "308 Peretsman Scully Hall"
    email         : "yk9446@princeton.edu"
    role:
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"

affiliation:
  - id            : "1"
    institution   : "Princeton University"

authornote: |
  Department of Psychology, Princeton University

abstract: |
  This is an example of the abstract. I took several code chunks from previous R markdown files that analyze child and parents' height.
  
  
keywords          : "height, genetics, parent-child relationship"
wordcount         : "000"

bibliography      : "r-references.bib"

floatsintext      : no
linenumbers       : no
draft             : yes
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no
numbersections    : yes

classoption       : "doc"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Methods {#sec:methods}

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis

We used `r cite_r("r-references.bib")` for all our analyses.

# Results


```{r}
library(ggplot2)
library(tidyverse)

child <- read.csv('child_height.csv')
parent <- read.csv('parent_height.csv')

child_parent_df <- child %>%
  left_join(parent, by = "family_id")

parent2child <- lm(child_ht~parent_ht, data = child_parent_df)
apa_model <- apa_print(parent2child)
```
The model showed `r apa_model$full_result`.
I actually do not know why the table appears under 'Data analysis' not 'Results' below.

(ref:table-caption) Regression statistics.

```{r}
apa_table(
  apa_model$table
)
```



(ref:figure-caption) child_ht~parent_ht.

```{r my-figure, fig.cap = "(ref:figure-caption)"}

ggplot(child_parent_df, aes(x = parent_ht, y = child_ht)) +
  geom_jitter(alpha = 0.3) 
```

Figure\ \@ref(fig:my-figure) shows the relationship between parent and child height.

# Discussion

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
